# Effects of Increased Compulsory Schooling Policy on Earnings

## Introduction

Compulsory schooling mandate of Turkey has been increased from 8 to 12 years in 2012. This policy change has impacted \<some_number\> of people when enacted. At the time, the policy was challenged by many families who argued that their children would not benefit from this additional years of education. They argued that their sons are better off working at a earlier age to gain *on-the-job* *training* which would earn them more money compared to additional schooling. Similarly these families argued that their daughters should marry and start a family instead of having education. While education and earnings have a multifaceted relationship, we investigate the causal aspects of this relationship using compulsory schooling policy change as an explanatory variable.

## Literature Review

The theory behind the possible impact of compulsory education on wages is human capital theory, which suggests that individuals increase their productivity by acquiring knowledge and skills through schooling and training (Schultz, 1961; Becker, 1962; Mincer, 1974).

Research describing the relationship between compulsory schooling and individual earnings is not very extensive. However, the relationship between these variables has been studied for a long time. It is possible to mention the work of Angrist and Krueger (1991), where they use compulsory schooling laws to estimate their causal effect on earnings. Furthermore, Harmon and Walker (1995) carried out a similar study, but based on the example of the United Kingdom, and obtained positive and significant results for the implementation of compulsory schooling. Subsequently, Oreopoulos (2006) extended the research beyond the United Kingdom and added to the analysis the cases of the United States, Canada and Ireland, obtaining similar results. Also worth mentioning is the study by Pischke and Von Wachter (2008), which examines the impact of an additional year of compulsory schooling on wages in Germany. However, the results are not consistent with those obtained in the previous studies, but the authors justify this with the specificities of the German case. Finally, Fang et al. (2012) also carried out the same exercise, but in China, and obtained more encouraging results than the other studies mentioned. However, this time the authors identify the impact of other aspects, in particular the rapid economic transformation and the increased demand for educated labour in the country.

## Data Cleaning and Formatting

```{r}

# Clear workspace
rm(list = ls())

# Import data using haven package for reading Stata files
library(haven)
library(dplyr)

# Read 2021 data
data_2021 <- read_dta("data/dta/data_2021.dta")

# Convert cinsiyet to numeric if it's not already
# Note: R handles this differently than Stata - we'll use as.numeric()
data_2021 <- data_2021 %>%
  mutate(
    cinsiyet = as.numeric(cinsiyet),
    year = 2021
  )

# Read 2022 data
data_2022 <- read_dta("data/dta/data_2022.dta")

# Convert cinsiyet to numeric for 2022 data
data_2022 <- data_2022 %>%
  mutate(
    cinsiyet = as.numeric(cinsiyet),
    year = 2022
  )

# Combine datasets
combined_data <- bind_rows(data_2022, data_2021)

# Create education-based treatment variables
combined_data <- combined_data %>%
  mutate(
    # Treatment 1: Higher than primary vs primary or less
    treatment1 = case_when(
      okul_biten_k %in% c(3, 41, 42, 511, 512, 52) ~ 1,
      okul_biten_k %in% c(1, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # Treatment 2: Higher than middle school vs middle school or less
    treatment2 = case_when(
      okul_biten_k %in% c(41, 42, 511, 512, 52) ~ 1,
      okul_biten_k %in% c(1, 2, 3) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # Treatment 3: Higher education vs high school or less
    treatment3 = case_when(
      okul_biten_k %in% c(511, 512, 52) ~ 1,
      okul_biten_k %in% c(1, 2, 3, 41, 42) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # Generate years of schooling
    yrs_school = case_when(
      okul_biten_k == 1 & okur_yazar == 1 ~ 2,
      okul_biten_k == 2 ~ 5,
      okul_biten_k == 3 ~ 8,
      okul_biten_k %in% c(41, 42) ~ 11,
      okul_biten_k %in% c(511, 512) ~ 15,
      okul_biten_k == 52 ~ 17,
      TRUE ~ NA_real_
    ),
    
    # Create gender dummy (1 if male)
    gender = as.numeric(cinsiyet == 1),
    
    # Convert wages to real terms (2021 as base year)
    # CPI2022/CPI2021 = 1.72
    gelir_gecenay_k = ifelse(year == 2022, 
                            gelir_gecenay_k/1.72, 
                            gelir_gecenay_k),
    
    # Calculate hourly wages
    hourly_wage = gelir_gecenay_k/(esas_fiili * 4.3),
    log_hourly_wage = log(hourly_wage)
  )

# Filter for wage and casual workers with positive earnings
# and working age population (18-65)
combined_data <- combined_data %>%
  filter(
    isteki_durum_k %in% c(11, 12),
    gelir_gecenay_k > 0,
    between(yas, 18, 65)
  )

# Save the prepared dataset
write_dta(combined_data, "data/dta/combined_data.dta")


# Convert haven_labelled variables to regular R formats
# This step is important because Stata's labelled variables are handled differently in R
combined_data <- combined_data %>%
  mutate(across(where(is.labelled), ~as.vector(.x)))

# The as.vector() function removes the Stata labels while preserving the values
# This makes the data more compatible with standard R functions

# Save as R's native RDS format
saveRDS(combined_data, "data/combined_data.rds")

```

## Directed Acyclic Graph (DAG)

We use DAG for plotting the associations between various co-variate and our outcome variable, earnings.

Outcome: Log hourly Earning = Y\
Explanatory Variables: Region, Gender, Age, Marital Status

```{r}

# Code for DAG showing relationships between earnings and demographic variables
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
library(dagitty)
library(ggdag)
library(gganimate)
library(ggthemes)
theme_set(theme_gray(base_size = 15))

# Set seed for reproducibility
set.seed(123)

# Create the DAG structure
# Y: Log hourly earnings
# R: Region
# G: Gender
# A: Age
# M: Marital Status
dag <- dagify(
    # Direct effects on Log hourly earnings (Y)
    Y ~ R + G + A + M,
    
    # Potential relationships between explanatory variables
    M ~ A + G,  # Marital status might be influenced by age and gender
    G ~ R,      # Gender composition might vary by region
    
    # Add variable labels for clarity
    labels = c(
        "Y" = "Log Hourly Earnings",
        "R" = "Region",
        "G" = "Gender",
        "A" = "Age",
        "M" = "Marital Status"
    )
) %>%
    tidy_dagitty()

# Create the visualization with improved styling
ggdag_classic(dag, node_size=10) + 
    theme_dag_blank() +
    ggtitle("DAG for Earnings Analysis") +
    theme(plot.title = element_text(hjust = 0.5))

```

## 

## Exploratory Data Analysis (EDA)

```{r}

# Add ggplot for explanatory variables and outcome variable

# 1. Age - 2 plots side by side - histogram for age frequency 

# 2. Marital Status - Pie Chart for each category 

# 3. Gender - Bar Chart with numbers

# 4. Region - Map of Turkey with each region annotated

# 5. Earnings - Histogram for earnings frequency

```

## Feature Engineering

```{r}

# Purpose: Create instrument variables and prepare control variables for analysis

# Load required libraries
library(haven)      # For reading Stata files
library(dplyr)      # For data manipulation
library(tidyr)      # For data cleaning
library(fastDummies) # For creating dummy variables


# Load the cleaned and combined data


################################################################################
# Create Instrument Variables 
################################################################################

# Create education reform exposure variables based on birth year
combined_data <- combined_data %>%
  mutate(
    # Treatment1: 1997 education reform (exposure if born after 1986)
    # This creates a binary indicator for those affected by the reform
    reform1997 = as.numeric(dogum_yil > 1986),
    
    # Treatment2: 2013 education reform (exposure if born after 2001)
    # Similar binary indicator for the 2013 reform
    reform2013 = as.numeric(dogum_yil > 2001),
    
    # Calculate university entry year (age 18) for each individual
    univ_entry_year = dogum_yil + 18
  )


################################################################################
# Control Variables
################################################################################

# Create region dummies
# In R, we'll first ensure ibbs_2 is properly factored, then create dummies
combined_data <- combined_data %>%
  mutate(
    region = as.factor(ibbs_2)
  )

# Create dummy variables for marital status
# We'll use dummy_cols from fastDummies package for this
combined_data <- dummy_cols(
  combined_data,
  select_columns = "medeni_durum",
  remove_selected_columns = FALSE,
  remove_first_dummy = TRUE
)

# Create birth year fixed effects
combined_data <- dummy_cols(
  combined_data,
  select_columns = "dogum_yil",
  remove_selected_columns = FALSE,
  remove_first_dummy = TRUE
)

# Create a list of control variables for use in models
# Note: In R, we'll create this differently than Stata's global macro
control_vars <- c(
  "region",              # Region fixed effects
  names(combined_data)[grep("^medeni_durum_", names(combined_data))],  # Marital status dummies
  "dogum_yil",          # Birth year
  names(combined_data)[grep("^dogum_yil_", names(combined_data))]      # Birth year dummies
)

# Save processed dataset if needed
saveRDS(combined_data, "data/analyzed_data.rds")

```

## Creating Training & Testing Data  

```{r}

# Purpose: Create Training/Testing data folds to be used by machine learning model

library(tidymodels)
library(tidyverse)
library(here)
library(readr)

turkey_data <- read_csv("Downloads/data.csv")

# Create a data split object, 
turkey_split <- initial_split(
  turkey_data, 
  prop = 0.75, 
  strata = log_hourly_wage
)

# Create the training & testing data
turkey_training <- training(turkey_split)
turkey_test <- testing(turkey_split)


# Distribution of selling_price in training data:
turkey_training %>% 
  summarize(min_sell_price = min(selling_price),
            max_sell_price = max(selling_price),
            mean_sell_price = mean(selling_price),
            sd_sell_price = sd(selling_price))

# Distribution of selling_price in test data
turkey_test %>% 
  summarize(min_sell_price = min(selling_price),
            max_sell_price = max(selling_price),
            mean_sell_price = mean(selling_price),
            sd_sell_price = sd(selling_price))
```

## References

Angrist, J. D., & Krueger, A. B. (1991). Does Compulsory School Attendance Affect Schooling and Earnings?. *The Quarterly Journal of Economics*, *106*(4), 979–1014. <https://doi.org/10.2307/2937954>

Becker, G. S. (1962). Investment in Human Capital: A Theoretical Analysis. *Journal of Political Economy*, *70*(5, Part 2), 9–49. <https://doi.org/10.1086/258724>

Fang, H., Eggleston, K. N., Rizzo, J. A., Rozelle, S., & Zeckhauser, R. J. (2012). *The Returns to Education in China: Evidence from the 1986 Compulsory Education Law* (Working Paper No. 18189). National Bureau of Economic Research. <https://doi.org/10.3386/w18189>

Harmon, C., & Walker, I. (1995). Estimates of the Economic Return to Schooling for the United Kingdom. *The American Economic Review*, *85*(5), 1278–1286.

Mincer, J. A. (1974). The Human Capital Earnings Function. In *Schooling, Experience, and Earnings* (pp. 83–96). NBER. <https://www.nber.org/books-and-chapters/schooling-experience-and-earnings/human-capital-earnings-function>

Oreopoulos, P. (2006). Estimating Average and Local Average Treatment Effects of Education when Compulsory Schooling Laws Really Matter. *American Economic Review*, *96*(1), 152–175. <https://doi.org/10.1257/000282806776157641>

Pischke, J.-S., & von Wachter, T. (2008). Zero Returns to Compulsory Schooling in Germany: Evidence and Interpretation. *The Review of Economics and Statistics*, *90*(3), 592–598. <https://doi.org/10.1162/rest.90.3.592>

Schultz, T. W. (1961). Investment in Human Capital. *The American Economic Review*, *51*(1), 1–17.
