# Effects of Increased Compulsory Schooling Policy on Earnings

## Introduction

Compulsory schooling mandate of Turkey has been increased from 8 to 12 years in 2012. This policy change has impacted \<some_number\> of people when enacted. At the time, the policy was challenged by many families who argued that their children would not benefit from this additional years of education. They argued that their sons are better off working at a earlier age to gain *on-the-job* *training* which would earn them more money compared to additional schooling. Similarly these families argued that their daughters should marry and start a family instead of having education. While education and earnings have a multifaceted relationship, we investigate the causal aspects of this relationship using compulsory schooling policy change as an explanatory variable.

## Literature Review

The theory behind the possible impact of compulsory education on wages is human capital theory, which suggests that individuals increase their productivity by acquiring knowledge and skills through schooling and training (Schultz, 1961; Becker, 1962; Mincer, 1974).

Research describing the relationship between compulsory schooling and individual earnings is not very extensive. However, the relationship between these variables has been studied for a long time. It is possible to mention the work of Angrist and Krueger (1991), where they use compulsory schooling laws to estimate their causal effect on earnings. Furthermore, Harmon and Walker (1995) carried out a similar study, but based on the example of the United Kingdom, and obtained positive and significant results for the implementation of compulsory schooling. Subsequently, Oreopoulos (2006) extended the research beyond the United Kingdom and added to the analysis the cases of the United States, Canada and Ireland, obtaining similar results. Also worth mentioning is the study by Pischke and Von Wachter (2008), which examines the impact of an additional year of compulsory schooling on wages in Germany. However, the results are not consistent with those obtained in the previous studies, but the authors justify this with the specificities of the German case. Finally, Fang et al. (2012) also carried out the same exercise, but in China, and obtained more encouraging results than the other studies mentioned. However, this time the authors identify the impact of other aspects, in particular the rapid economic transformation and the increased demand for educated labour in the country.

## Data Cleaning

1.  Combine 2021 - 2022 Datasets

```{r}

# Read dta data
data_2021 <- read_dta("data/dta/data_2021.dta")
data_2022 <- read_dta("data/dta/data_2022.dta")

# Combine datasets
combined_data <- bind_rows(data_2022, data_2021)

# Save it as RDS
saveRDS(combined_data, "data/combined_data.rds")

```

2.  Rename our variables to English

```{r}

# first rename the variables: 
# yas->age, medeni_durum->marital_status, ibbs_2->region, cinsiyet->gender
...


```

## Feature Engineering

1.  Encode some continues variables to categorical

```{r}

# now make the marital_status, region and age categorical variables
...


```

2.  Create log_hourly_wage and treatment variables for compulsory schooling

```{r}

# now create log_hourly_wage variable
...

# then lets create treatment variables for individuals effected by compulsory schooling policy
...


```

## Exploratory Data Analysis (EDA)

1.  Lets see the distribution of log hourly wages

```{r}

# ggplot for historgram



```

2.  Lets see the how many people are effected by compulsory schooling policy change

```{r}

# now create log_hourly_wage variable
...



```

## 

## Directed Acyclic Graph (DAG)

We use DAG for plotting the associations between various co-variate and our outcome variable, earnings.

Outcome: Log hourly Earning = Y\
Explanatory Variables: Region, Gender, Age, Marital Status

```{r}

# Code for DAG showing relationships between earnings and demographic variables
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
library(dagitty)
library(ggdag)
library(gganimate)
library(ggthemes)
theme_set(theme_gray(base_size = 15))

# Set seed for reproducibility
set.seed(123)

# Create the DAG structure
# Y: Log hourly earnings
# R: Region
# G: Gender
# A: Age
# M: Marital Status
dag <- dagify(
    # Direct effects on Log hourly earnings (Y)
    Y ~ R + G + A + M,
    
    # Potential relationships between explanatory variables
    M ~ A + G,  # Marital status might be influenced by age and gender
    G ~ R,      # Gender composition might vary by region
    
    # Add variable labels for clarity
    labels = c(
        "Y" = "Log Hourly Earnings",
        "R" = "Region",
        "G" = "Gender",
        "A" = "Age",
        "M" = "Marital Status"
    )
) %>%
    tidy_dagitty()

# Create the visualization with improved styling
ggdag_classic(dag, node_size=10) + 
    theme_dag_blank() +
    ggtitle("DAG for Earnings Analysis") +
    theme(plot.title = element_text(hjust = 0.5))

```

## 

## 

```{r}

# Purpose: Create instrument variables and prepare control variables for analysis

# Load required libraries
library(haven)      # For reading Stata files
library(dplyr)      # For data manipulation
library(tidyr)      # For data cleaning
library(fastDummies) # For creating dummy variables


# Load the cleaned and combined data


################################################################################
# Create Policy Variables
################################################################################

# Create education reform exposure variables based on birth year
combined_data <- combined_data %>%
  mutate(
    # Treatment1: 1997 education reform (exposure if born after 1986)
    # This creates a binary indicator for those affected by the reform
    reform1997 = as.numeric(dogum_yil > 1986),
    
    # Treatment2: 2013 education reform (exposure if born after 2001)
    # Similar binary indicator for the 2013 reform
    reform2013 = as.numeric(dogum_yil > 2001),
    
    # Calculate university entry year (age 18) for each individual
    univ_entry_year = dogum_yil + 18
  )


################################################################################
# Control Variables
################################################################################

# Create region dummies
# In R, we'll first ensure ibbs_2 is properly factored, then create dummies
combined_data <- combined_data %>%
  mutate(
    region = as.factor(ibbs_2)
  )

# Create dummy variables for marital status
# We'll use dummy_cols from fastDummies package for this
combined_data <- dummy_cols(
  combined_data,
  select_columns = "medeni_durum",
  remove_selected_columns = FALSE,
  remove_first_dummy = TRUE
)

# Create birth year fixed effects
combined_data <- dummy_cols(
  combined_data,
  select_columns = "dogum_yil",
  remove_selected_columns = FALSE,
  remove_first_dummy = TRUE
)

# Create a list of control variables for use in models
# Note: In R, we'll create this differently than Stata's global macro
control_vars <- c(
  "region",              # Region fixed effects
  names(combined_data)[grep("^medeni_durum_", names(combined_data))],  # Marital status dummies
  "dogum_yil",          # Birth year
  names(combined_data)[grep("^dogum_yil_", names(combined_data))]      # Birth year dummies
)

# Save processed dataset if needed
saveRDS(combined_data, "data/analyzed_data.rds")

```

## Splitting Training & Testing Data  

```{r}

# Purpose: Create Training/Testing data folds to be used by machine learning model

library(tidymodels)
library(tidyverse)
library(here)
library(readr)

turkey_data <- read_csv("Downloads/data.csv")

# Create a data split object, 
turkey_split <- initial_split(
  turkey_data, 
  prop = 0.75, 
  strata = log_hourly_wage
)

# Create the training & testing data
turkey_training <- training(turkey_split)
turkey_test <- testing(turkey_split)


# Distribution of selling_price in training data:
turkey_training %>% 
  summarize(min_sell_price = min(selling_price),
            max_sell_price = max(selling_price),
            mean_sell_price = mean(selling_price),
            sd_sell_price = sd(selling_price))

# Distribution of selling_price in test data
turkey_test %>% 
  summarize(min_sell_price = min(selling_price),
            max_sell_price = max(selling_price),
            mean_sell_price = mean(selling_price),
            sd_sell_price = sd(selling_price))
```

## Training Machine Learning Models

In this section, we will employ both regression and classification models and measure their performances.\
This process will allow us to find out the best performing model so that we can improve on that.

### Classification Model

First we employ a binary classification model for predicting weather some is *employed* or *unemployed* \
This

```{r}

# First

# Save processed dataset if needed
saveRDS(combined_data, "data/analyzed_data.rds")

```

#### Performance Measurements

```{r}

# First

# Save processed dataset if needed
saveRDS(combined_data, "data/analyzed_data.rds")

```

### Regression Model

Now we employ linear & polynomial regression models for predicting the log hour earning of a employed individual.

```{r}

# First

# Save processed dataset if needed
saveRDS(combined_data, "data/analyzed_data.rds")

```

#### Performance Measurements

```{r}

# First

# Save processed dataset if needed
saveRDS(combined_data, "data/analyzed_data.rds")

```

### 

## Conclusion

## References

Angrist, J. D., & Krueger, A. B. (1991). Does Compulsory School Attendance Affect Schooling and Earnings?. *The Quarterly Journal of Economics*, *106*(4), 979–1014. <https://doi.org/10.2307/2937954>

Becker, G. S. (1962). Investment in Human Capital: A Theoretical Analysis. *Journal of Political Economy*, *70*(5, Part 2), 9–49. <https://doi.org/10.1086/258724>

Fang, H., Eggleston, K. N., Rizzo, J. A., Rozelle, S., & Zeckhauser, R. J. (2012). *The Returns to Education in China: Evidence from the 1986 Compulsory Education Law* (Working Paper No. 18189). National Bureau of Economic Research. <https://doi.org/10.3386/w18189>

Harmon, C., & Walker, I. (1995). Estimates of the Economic Return to Schooling for the United Kingdom. *The American Economic Review*, *85*(5), 1278–1286.

Mincer, J. A. (1974). The Human Capital Earnings Function. In *Schooling, Experience, and Earnings* (pp. 83–96). NBER. <https://www.nber.org/books-and-chapters/schooling-experience-and-earnings/human-capital-earnings-function>

Oreopoulos, P. (2006). Estimating Average and Local Average Treatment Effects of Education when Compulsory Schooling Laws Really Matter. *American Economic Review*, *96*(1), 152–175. <https://doi.org/10.1257/000282806776157641>

Pischke, J.-S., & von Wachter, T. (2008). Zero Returns to Compulsory Schooling in Germany: Evidence and Interpretation. *The Review of Economics and Statistics*, *90*(3), 592–598. <https://doi.org/10.1162/rest.90.3.592>

Schultz, T. W. (1961). Investment in Human Capital. *The American Economic Review*, *51*(1), 1–17.
